---
title: "CodeBook.md"
output: html_document
---

## Getting and Cleaning Data Course Project
## Peer-graded Assignment

# Data
This project assignment uses data downloaded from:https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

The task of the assignment is to create one R script called run_analysis.R that does the following:  
- 1. Merges the training and the test sets to create one data set.  
- 2. Extracts only the measurements on the mean and standard deviation for each measurement.   
- 3. Uses descriptive activity names to name the activities in the data set.  
- 4. Appropriately labels the data set with descriptive variable names.  
- 5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.  

# Downloading the files

Check present working directory before downloading files.We then unzip the files.

```{r}
getwd()
zippedfile <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(zippedfile, destfile = "./data/Dataset.zip", method= "curl")
list.files("./data")
unzip(zipfile ="./data/Dataset.zip", exdir = "./data")
```

The folders and files are renamed by removing spaces and changing to lowercase.

```{r message = FALSE, warning=FALSE}
recursive_replace_lowercase<-function(path=".", replace=" ", with="_",
                                      lowercase=TRUE) {
        # this is the base case
        filelist<-list.files(path, full.names=TRUE)
        if (lowercase) {
                for(filename in filelist)
                        file.rename(filename,gsub(replace,with,tolower(filename)))
        } else {
                for(filename in filelist)
                        file.rename(filename,gsub(replace,with,filename))
        }
        # and this is the recursive part
        dirlist<-list.dirs(path, full.names=TRUE, recursive=FALSE)
        if(length(dirlist)) {
                for(dirname in dirlist)
                        recursive_replace_lowercase(dirname, replace=replace, with=with,
                                                    lowercase=lowercase)
        }
}
recursive_replace_lowercase()
```

Get a list of files & directories in the uci_har_dataset directory:
```{r}
list.files("./data/uci_har_dataset/")
```

# Description of the data
As described in the features_info.txt file provided for the assignment, the features are from the accelerometer and gyroscope 3-axial raw signals tAccelerometer-XYZ and tGyroscope-XYZ. 

# Units of data
Data with prefix 't' (to denote time) are time domain signals, captured at a constant rate of 50 Hz, then filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. The acceleration signal was then separated into body and gravity acceleration signals (TimeBodyAccelerometer-XYZ and TimeGravityAccelerometer-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

The body linear acceleration and angular velocity were later derived in time to obtain Jerk signals (TimeBodyAccelerometerJerk-XYZ and TimeBodyGyroscopeJerk-XYZ). The magnitude of these 3-D signals were calculated using the Euclidean norm (TimeBodyAccelerometerMagnitude, TimeGravityAccelerometerMagnitude, TimeBodyAccelerometerJerkMagnitude, TimeBodyGyroscopeMagnitude, TimeBodyGyroscopeJerkMagnitude). 

Data with prefix 'f' are frequency domain signals. AFast Fourier Transform (FFT) was applied to some of these signals to get FrequencyBodyAccelerometer-XYZ, FrequencyBodyAccelerometerJerk-XYZ, FrequencyBodyGyroscope-XYZ, FrequencyBodyAccelerometerJerkMagnitude, FrequencyBodyGyroscopeMagnitude, FrequencyBodyGyroscopeJerkMagnitude.

These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.


TimeBodyAccelerometer-XYZ  
TimeGravityAccelerometer-XYZ  
TimeBodyAccelerometerJerk-XYZ  
TimeBodyGyroscope-XYZ  
TimeBodyGyroscopeJerk-XYZ  
TimeBodyAccelerometerMagnitude  
TimeGravityAccelerometerMagnitude  
TimeBodyAccelerometerJerkMagnitude  
TimeBodyGyroscopeMagnitude  
TimeBodyGyroscopeJerkMagnitude  
FrequencyBodyAccelerometer-XYZ  
FrequencyBodyAccelerometerJerk-XYZ  
FrequencyBodyGyroscope-XYZ  
FrequencyBodyAccelerometerMagnitude  
FrequencyBodyAccelerometerJerkMagnitude  
FrequencyBodyGyroscopeMagnitude  
FrequencyBodyGyroscopeJerkMagnitude  

The final set of variables that were selected from these signals are: 
Mean: Mean value
STD: Standard deviation

Additional vectors obtained by averaging the signals in a signal window sample. These are used on the angle() variable:

gravityMean
TimeBodyAccelerometerMean
TimeBodyAccelerometerJerkMean
TimeBodyGyroscopeMean
TimeBodyGyroscopeJerkMean

The complete list of variables of each feature vector is available in 'features.txt' 

# Subjects
Subjects are numbered from 1 to 30

# Activity Labels
Tthe experiment monitored via a smartphone the six different activities:

activity_label           activity  
             1            WALKING  
             2   WALKING_UPSTAIRS  
             3 WALKING_DOWNSTAIRS  
             4            SITTING  
             5           STANDING  
             6             LAYING  



# 1A. Read in the files describing the activity_labels and features.

Read the activity_labels.txt file as activity_labels and name the columns:

```{r}
activity_labels <- read.table("./data/uci_har_dataset/activity_labels.txt")
colnames(activity_labels, do.NULL = FALSE)
colnames(activity_labels) <- c("activity_label","activity")
head(activity_labels)
```

Read the features.txt" file and mae as features and name the columns:
```{r}
features<- read.table("./data/uci_har_dataset/features.txt")
colnames(features, do.NULL = FALSE)
colnames(features) <- c("feature_no","feature")
head(features)
```

# 1B. Read in the training set files and merge the files by using cbind

Get a list of files in the train directory:

```{r}
list.files("./data/uci_har_dataset/train")
```

Read in the subject_train.txt file as train_subjects and label the column name as "subject":
```{r}
train_subjects<- read.table("./data/uci_har_dataset/train/subject_train.txt")
colnames(train_subjects, do.NULL = FALSE)
colnames(train_subjects) <- c("subject")
head(train_subjects)
```

Read in the y_train.txt file as train_label and label the column name as "activity_label":
```{r}
train_label<- read.table("./data/uci_har_dataset/train/y_train.txt")
colnames(train_label, do.NULL = FALSE)
colnames(train_label) <- c("activity_label")
head(train_label)
```

Read in the x_train.txt file as train_set, and name the columnns by using the 2nd column in the "features" file:
```{r}
train_set <- read.table("./data/uci_har_dataset/train/x_train.txt")
train_set_colnames <- features[,2]
colnames(train_set, do.NULL = FALSE)
colnames(train_set) <- train_set_colnames
head(train_set)[1:5,1:5]
```

Merge the train_subjects, train_label and train_set data by using cbind and check the features of the merged traindf dataframe
```{r}
traindf <- cbind(train_subjects,train_label,train_set)
dim(traindf)
head(traindf)[1:5,1:5]
```


# 1C. Read in the test set files and merge the files by using cbind

Get a list of files in the test directory:
```{r}
list.files("./data/uci_har_dataset/test")
```

Read in the subject_test.txt file as train_subjects and label the column name as "subject":
```{r}
test_subjects<- read.table("./data/uci_har_dataset/test/subject_test.txt")
colnames(test_subjects, do.NULL = FALSE)
colnames(test_subjects) <- c("subject")
head(test_subjects)
```

Read in the test label file and name the column as "activity_label"
```{r}
test_label<- read.table("./data/uci_har_dataset/test/y_test.txt")
colnames(test_label, do.NULL = FALSE)
colnames(test_label) <- c("activity_label")
head(test_label)
```

Read in the x_train.txt file as test_set
```{r}
test_set <- read.table("./data/uci_har_dataset/test/x_test.txt")
test_set_colnames <- features[,2]
colnames(test_set, do.NULL = FALSE)
colnames(test_set) <- test_set_colnames
head(test_set)[1:5,1:5]
```

Merge the test_subjects, test_label and test_set data by using cbind and check the features of the merged testdf dataframe.
```{r}
testdf <- cbind(test_subjects, test_label, test_set)
head(testdf)[1:6,1:6]
dim(testdf)
```


# 1D. Merge the test and training sets by using rbind and check the features of the merged dataframe.
```{r}
mergedf <- rbind(traindf, testdf)
dim(mergedf)
head(mergedf)[1:5,1:5]
```


# 2. Extracts only the measurements on the mean and standard deviation for each measurement from the mered dataframe.

Load the dplyr package
```{r message = FALSE, warning=FALSE}
library(dplyr)
```

Select the variables containing mean and standard deviation using yje select function:
```{r}
meanstddf <- select(mergedf, 1:3, contains("mean"), contains("std"))
dim(meanstddf)
head(meanstddf)[1:5,1:5]
```


# 3. Uses descriptive activity names to name the activities in the data set

Use "left-join" to merge the meanstddf and activity_labels dataframes by "activity_label". The "activity_labels" data was read in at part 1A. 
```{r}
meanstddf2 <- left_join(meanstddf, activity_labels, "activity_label")
head(meanstddf2)[1:5,1:5]
dim(meanstddf2)
```

Rearrange the columns, replace "activity_label" with the "activity" column by moving it to the second column, assign dataframe
```{r}
meanstddf3 <- meanstddf2[c(1, 89, 3:88)]
head(meanstddf3)[1:5,1:5]
dim(meanstddf3)
```

# 4.Appropriately labels the data set with descriptive variable names.

Load the stringr package
```{r}
library(stringr)
```


Replace the shortened words in the variable names with actual words: BodyBody, with Body, Acc with Accelerometer, Gyro with Gyroscope, Mag with Magnitude, t with Time, f and -freq() with Frequency, tBody with TimeBody, angle with Angle, and gravity with Gravity, -mean() with Mean, and -std() with STD:
```{r}
names(meanstddf3) <- gsub("Acc", "Accelerometer",names(meanstddf3)) %>% 
        str_replace_all("Gyro", "Gyroscope") %>% 
        str_replace_all("BodyBody", "Body") %>%
        str_replace_all("Mag", "Magnitude") %>% 
        str_replace_all("^t", "Time") %>% 
        str_replace_all("^f", "Frequency") %>% 
        str_replace_all("tBody", "TimeBody") %>% 
        str_replace_all("-mean\\(\\)", "Mean") %>% 
        str_replace_all("-std\\(\\)", "STD") %>% 
        str_replace_all("-freq\\(\\)", "Frequency") %>% 
        str_replace_all("\\(\\)","") %>% 
        str_replace_all("angle", "Angle") %>% 
        str_replace_all("gravity", "Gravity") %>%
        print
```

#check the features of the new meanstddf3 dataframe:
```{r}
dim(meanstddf3)
head(meanstddf3)[1:5,1:5]
```

# 5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

Use dplyr to group by subject and activity
```{r}
mean_activity<- meanstddf3 %>% 
        group_by(subject,activity) %>% 
        # summarise accross the variables that are numeric by mean, 
        # prepend the label "mean_" before each column
        summarise(across(where(is.numeric), ~ mean(.x), .names = "mean_{.col}")) %>% 
        print
```

Create the tidy data set as a csv file
```{r}
write.table(mean_activity, "mean_activity.txt",row.name=FALSE)
```


