---
title: "CodeBook.md"
output: html_document
---

## Getting and Cleaning Data Course Project
The task of the assignment is to create one R script called run_analysis.R that does the following:  
- 1. Merges the training and the test sets to create one data set.  
- 2. Extracts only the measurements on the mean and standard deviation for each measurement.   
- 3. Uses descriptive activity names to name the activities in the data set.  
- 4. Appropriately labels the data set with descriptive variable names.  
- 5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.  

# Data
This project assignment uses data from: https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip


# Description of the data
Data features collected from the smartphone accelerometer and gyroscope are the 3-axial raw signals tAccelerometer-XYZ and tGyroscope-XYZ. 

# Units of data
Data with prefix 't' are time domain signals, captured at a constant rate of 50 Hz, then filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. The acceleration signal was then separated into body and gravity acceleration signals (TimeBodyAccelerometer-XYZ and TimeGravityAccelerometer-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

The body linear acceleration and angular velocity were later derived in time to obtain Jerk signals (TimeBodyAccelerometerJerk-XYZ and TimeBodyGyroscopeJerk-XYZ). The magnitude of these 3-D signals were calculated using the Euclidean norm (TimeBodyAccelerometerMagnitude, TimeGravityAccelerometerMagnitude, TimeBodyAccelerometerJerkMagnitude, TimeBodyGyroscopeMagnitude, TimeBodyGyroscopeJerkMagnitude). 

Data with prefix 'f' are frequency domain signals. A Fast Fourier Transform (FFT) was applied to some of these signals to get FrequencyBodyAccelerometer-XYZ, FrequencyBodyAccelerometerJerk-XYZ, FrequencyBodyGyroscope-XYZ, FrequencyBodyAccelerometerJerkMagnitude, FrequencyBodyGyroscopeMagnitude, and FrequencyBodyGyroscopeJerkMagnitude.

These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote the 3-axial signals in the X, Y and Z directions.

TimeBodyAccelerometer-XYZ  
TimeGravityAccelerometer-XYZ  
TimeBodyAccelerometerJerk-XYZ  
TimeBodyGyroscope-XYZ  
TimeBodyGyroscopeJerk-XYZ  
TimeBodyAccelerometerMagnitude  
TimeGravityAccelerometerMagnitude  
TimeBodyAccelerometerJerkMagnitude  
TimeBodyGyroscopeMagnitude  
TimeBodyGyroscopeJerkMagnitude  
FrequencyBodyAccelerometer-XYZ  
FrequencyBodyAccelerometerJerk-XYZ  
FrequencyBodyGyroscope-XYZ  
FrequencyBodyAccelerometerMagnitude  
FrequencyBodyAccelerometerJerkMagnitude  
FrequencyBodyGyroscopeMagnitude  
FrequencyBodyGyroscopeJerkMagnitude  

Additional vectors obtained by averaging the signals in a signal window sample. These are used on the Angle variable:

gravityMean
TimeBodyAccelerometerMean
TimeBodyAccelerometerJerkMean
TimeBodyGyroscopeMean
TimeBodyGyroscopeJerkMean

The final set of variables that were selected from these signals are **Mean** and **Standard deviation (STD)** values. The complete list of variables of each feature vector is available in 'features.txt' 

# Subjects
Subjects are numbered from 1 to 30

# Activity Labels
The experiment monitored six different activities: 

activity_label           activity  
             1            WALKING  
             2   WALKING_UPSTAIRS  
             3 WALKING_DOWNSTAIRS  
             4            SITTING  
             5           STANDING  
             6             LAYING  



# Downloading the files

Check present working directory before downloading files, then unzip the files.
```{r}
getwd()
zippedfile <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(zippedfile, destfile = "./data/Dataset.zip", method= "curl")
list.files("./data")
unzip(zipfile ="./data/Dataset.zip", exdir = "./data")
```

Rename the folders and files by replacing the spaces with "_" and changing to lowercase:  
```{r message = FALSE, warning=FALSE}
recursive_replace_lowercase<-function(path=".", replace=" ", with="_",
                                      lowercase=TRUE) {
        # this is the base case
        filelist<-list.files(path, full.names=TRUE)
        if (lowercase) {
                for(filename in filelist)
                        file.rename(filename,gsub(replace,with,tolower(filename)))
        } else {
                for(filename in filelist)
                        file.rename(filename,gsub(replace,with,filename))
        }
        # and this is the recursive part
        dirlist<-list.dirs(path, full.names=TRUE, recursive=FALSE)
        if(length(dirlist)) {
                for(dirname in dirlist)
                        recursive_replace_lowercase(dirname, replace=replace, with=with,
                                                    lowercase=lowercase)
        }
}
recursive_replace_lowercase()
```

Get a list of files & directories in the uci_har_dataset directory:
```{r}
list.files("./data/uci_har_dataset/")
```


# 1A. Read in the files describing the activity_labels and features.

Read the activity_labels.txt file as activity_labels and name the columns:
```{r}
activity_labels <- read.table("./data/uci_har_dataset/activity_labels.txt")
colnames(activity_labels) <- c("activity_label","activity")
```

Read the features.txt" file and mae as features and name the columns:
```{r}
features<- read.table("./data/uci_har_dataset/features.txt")
```

# 1B. Read in the training set files and merge the files by using cbind

Read in the training set files
```{r}
train_subjects<- read.table("./data/uci_har_dataset/train/subject_train.txt")
train_label<- read.table("./data/uci_har_dataset/train/y_train.txt")
train_set <- read.table("./data/uci_har_dataset/train/x_train.txt")
```

Merge the train_subjects, train_label and train_set data by using cbind:
```{r}
traindf <- cbind(train_subjects,train_label,train_set)
```


# 1C. Read in the test set files and merge the files by using cbind

Read in the test set files:
```{r}
test_subjects<- read.table("./data/uci_har_dataset/test/subject_test.txt")
test_label<- read.table("./data/uci_har_dataset/test/y_test.txt")
test_set <- read.table("./data/uci_har_dataset/test/x_test.txt")
```

Merge the test_subjects, test_label and test_set data by using cbind:
```{r}
testdf <- cbind(test_subjects, test_label, test_set)
```

# 1D. Merge the test and training sets by using rbind.
```{r}
mergedf <- rbind(traindf, testdf)
```

# 1E. Name the columns of the dataframe, using the variables description in the features file:
```{r}
train_set_colnames <- features[,2]
colnames(mergedf) <- c("subject", "activity_label", train_set_colnames)
str(mergedf)
head(mergedf)[,1:5]
```

# 2. Extracts only the measurements on the mean and standard deviation for each measurement from the mered dataframe.

Load the "dplyr" package
```{r message = FALSE, warning=FALSE}
library(dplyr)
```

Select the variables containing mean and standard deviation using the "select" function:
```{r}
meandf <- select(mergedf, 1:2, matches("[Mm]ean"), contains("std"))
dim(meandf)
head(meandf)[,1:5]
```


# 3. Uses descriptive activity names to name the activities in the data set

Use "left-join" to merge the meandf and activity_labels dataframes joining by "activity_label":
```{r}
meandf2 <- left_join(meandf,activity_labels, "activity_label")
dim(meandf2)
head(meandf2)[,85:89]
```

Rearrange the columns: moving "activity_label" to column 2 and assign dataframe to meandf3:
```{r}
meandf3 <- meandf2[c(1, 89, 3:88)]
head(meandf3)[,1:5]
dim(meandf3)
```

# 4.Appropriately labels the data set with descriptive variable names.

Load the "stringr" package.
```{r}
library(stringr)
```

Replace abbreviations in variable names with actual words: BodyBody, with Body; Acc with Accelerometer; Gyro with Gyroscope; Mag with Magnitude; t with Time; f and -freq() with Frequency; tBody with TimeBody; angle with Angle; gravity with Gravity, -mean() with Mean, and -std() with STD:
```{r}
names(meandf3) <- gsub("Acc", "Accelerometer",names(meandf3)) %>% 
        str_replace_all("Gyro", "Gyroscope") %>% 
        str_replace_all("BodyBody", "Body") %>%
        str_replace_all("Mag", "Magnitude") %>% 
        str_replace_all("^t", "Time") %>% 
        str_replace_all("^f", "Frequency") %>% 
        str_replace_all("tBody", "TimeBody") %>%
        str_replace_all("\\(","") %>%
        str_replace_all("\\)","") %>%
        str_replace_all("\\-", "") %>%
        str_replace_all("\\[Mm]ean", "Mean") %>% 
        str_replace_all("std", "STD") %>% 
        str_replace_all("freq", "Frequency") %>% 
        str_replace_all("angle", "Angle") %>% 
        str_replace_all("gravity", "Gravity") %>%
        print
```

Check the features of the new meanstddf3 dataframe:
```{r}
dim(meandf3)
head(meandf3)[,1:10]
```

# 5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

Use dplyr to group by subject and activity
```{r}
mean_activity<- meandf3 %>% 
        group_by(subject,activity) %>% 
        # summarise accross the variables that are numeric by mean, 
        # prepend the label "mean_" before each column
        summarise(across(where(is.numeric), ~ mean(.x), .names = "mean_{.col}")) %>% 
        print
```

Create the tidy data set as a csv file
```{r}
write.table(mean_activity, "mean_activity.txt",row.name=FALSE)
```


