---
title: "README"
author: "stan"
date: "12/27/2020"
output: html_document
---

# Peer-graded Assignment: Getting and Cleaning Data Course Project

This repository contains the following files:

 - This README.rmd file
 - An R script called run_analysis.R
 - A codebook.rmd file that describes the variables, the data, and any transformations and work performed to clean up the data.
 - A mean_activity.csv file containing the tidy data set.
 
The link to the Github repository with the script for performing the analysis and the listed files are as follows: https://github.com/shyhhantan/DataClean

# Instructions

This project assignment uses data downloaded from:https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

The task of the assignment is to create one R script called run_analysis.R that does the following:  
- 1. Merges the training and the test sets to create one data set.  
- 2. Extracts only the measurements on the mean and standard deviation for each measurement.   
- 3. Uses descriptive activity names to name the activities in the data set.  
- 4. Appropriately labels the data set with descriptive variable names.  
- 5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.  

# The steps performed are described in the CodeBook:

# Downloading the data and unzipping the files:

The files are downloaded from: https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

and unzipped.

# To read in the features, training data, test data are 
# 1A. Read in the files describing the activity_labels and features.

Read the activity_labels.txt file as activity_labels and name the columns:

```{r}
activity_labels <- read.table("./data/uci_har_dataset/activity_labels.txt")
colnames(activity_labels) <- c("activity_label","activity")
```

Read the features.txt" file and mae as features and name the columns:
```{r}
features<- read.table("./data/uci_har_dataset/features.txt")
colnames(features) <- c("feature_no","feature")
```

# 1B. Read in the training set files and merge the files by using cbind

Read in the subject_train.txt file as train_subjects and label the column name as "subject":
```{r}
train_subjects<- read.table("./data/uci_har_dataset/train/subject_train.txt")
colnames(train_subjects) <- c("subject")
```

Read in the y_train.txt file as train_label and label the column name as "activity_label":
```{r}
train_label<- read.table("./data/uci_har_dataset/train/y_train.txt")
colnames(train_label) <- c("activity_label")
```

Read in the x_train.txt file as train_set, and name the columnns by using the 2nd column in the "features" file:
```{r}
train_set <- read.table("./data/uci_har_dataset/train/x_train.txt")
train_set_colnames <- features[,2]
colnames(train_set) <- train_set_colnames
head(train_set)[1:5,1:5]
```

Merge the train_subjects, train_label and train_set data by using cbind and check the features of the merged traindf dataframe
```{r}
traindf <- cbind(train_subjects,train_label,train_set)
dim(traindf)
head(traindf)[1:5,1:5]
```


# 1C. Read in the test set files and merge the files by using cbind

Read in the subject_test.txt file as train_subjects and label the column name as "subject":
```{r}
test_subjects<- read.table("./data/uci_har_dataset/test/subject_test.txt")
colnames(test_subjects) <- c("subject")
head(test_subjects)
```

Read in the test label file and name the column as "activity_label"
```{r}
test_label<- read.table("./data/uci_har_dataset/test/y_test.txt")
colnames(test_label) <- c("activity_label")
head(test_label)
```

Read in the x_train.txt file as test_set
```{r}
test_set <- read.table("./data/uci_har_dataset/test/x_test.txt")
test_set_colnames <- features[,2]
colnames(test_set) <- test_set_colnames
head(test_set)[1:5,1:5]
```

Merge the test_subjects, test_label and test_set data by using cbind and check the features of the merged testdf dataframe.
```{r}
testdf <- cbind(test_subjects, test_label, test_set)
head(testdf)[1:6,1:6]
dim(testdf)
```

# 1D. Merge the test and training sets by using rbind and check the features of the merged dataframe.
```{r}
mergedf <- rbind(traindf, testdf)
dim(mergedf)
head(mergedf)[1:5,1:5]
```


# 2. Extracts only the measurements on the mean and standard deviation for each measurement from the mered dataframe.

Load the dplyr package
```{r message = FALSE, warning=FALSE}
library(dplyr)
```

Select the variables containing mean and standard deviation using yje select function:
```{r}
meanstddf <- select(mergedf, 1:3, contains("mean"), contains("std"))
dim(meanstddf)
head(meanstddf)[1:5,1:5]
```


# 3. Uses descriptive activity names to name the activities in the data set

Use "left-join" to merge the meanstddf and activity_labels dataframes by "activity_label". The "activity_labels" data was read in at part 1A. 
```{r}
meanstddf2 <- left_join(meanstddf, activity_labels, "activity_label")
head(meanstddf2)[1:5,1:5]
dim(meanstddf2)
```

Rearrange the columns, replace "activity_label" with the "activity" column by moving it to the second column, assign dataframe
```{r}
meanstddf3 <- meanstddf2[c(1, 89, 3:88)]
head(meanstddf3)[1:5,1:5]
dim(meanstddf3)
```

# 4.Appropriately labels the data set with descriptive variable names.

Load the stringr package
```{r}
library(stringr)
```

Replace the shortened words in the variable names with actual words: BodyBody, with Body, Acc with Accelerometer, Gyro with Gyroscope, Mag with Magnitude, t with Time, f and -freq() with Frequency, tBody with TimeBody, angle with Angle, and gravity with Gravity, -mean() with Mean, and -std() with STD:
```{r}
names(meanstddf3) <- gsub("Acc", "Accelerometer",names(meanstddf3)) %>% 
        str_replace_all("Gyro", "Gyroscope") %>% 
        str_replace_all("BodyBody", "Body") %>%
        str_replace_all("Mag", "Magnitude") %>% 
        str_replace_all("^t", "Time") %>% 
        str_replace_all("^f", "Frequency") %>% 
        str_replace_all("tBody", "TimeBody") %>% 
        str_replace_all("-mean\\(\\)", "Mean") %>% 
        str_replace_all("-std\\(\\)", "STD") %>% 
        str_replace_all("-freq\\(\\)", "Frequency") %>% 
        str_replace_all("\\(\\)","") %>% 
        str_replace_all("angle", "Angle") %>% 
        str_replace_all("gravity", "Gravity") %>%
        print
```

#check the features of the new meanstddf3 dataframe:
```{r}
dim(meanstddf3)
head(meanstddf3)[1:5,1:5]
```

# 5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

Use dplyr to group by subject and activity
```{r}
mean_activity<- meanstddf3 %>% 
        group_by(subject,activity) %>% 
        # summarise accross the variables that are numeric by mean, 
        # prepend the label "mean_" before each column
        summarise(across(where(is.numeric), ~ mean(.x), .names = "mean_{.col}")) %>% 
        print
```

Create the tidy data set as a csv file
```{r}
write.table(mean_activity, "mean_activity.txt",row.name=FALSE)
```


